---
title: "Airbnb prices across European cities"
author: "Jane Shadrina"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

This analysis studies Airbnb prices across European cities using listings,
neighbourhood boundaries, and tourist amenities from OpenStreetMap. Listings
are nested within neighbourhoods and cities.

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  dpi = 300
)

options(
  scipen = 999,
  dplyr.summarise.inform = FALSE
)
```

```{r packages}
required_packages <- c(
  "tidyverse",
  "sf",
  "osmdata",
  "lme4",
  "lmerTest",
  "modelsummary",
  "ggeffects",
  "cowplot"
)

missing_packages <- setdiff(required_packages, rownames(installed.packages()))

if (length(missing_packages) > 0) {
  stop(
    "Install the missing packages before running the analysis: ",
    paste(missing_packages, collapse = ", ")
  )
}

invisible(lapply(required_packages, library, character.only = TRUE))
sf::sf_use_s2(FALSE)
```

## Configuration

### Paths

The data directory can be set with `AIRBNB_DATA_DIR`. Generated files stay in
the local `outputs` directory.

```{r paths}
data_dir <- Sys.getenv(
  "AIRBNB_DATA_DIR",
  unset = file.path("..", "multievel database")
)
output_dir <- "outputs"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
```

### Cities

```{r cities}
all_cities <- c(
  "london",
  "madrid",
  "amsterdam",
  "barcelona",
  "berlin",
  "lisbon",
  "paris",
  "rome",
  "venice",
  "vienna"
)

analytic_cities <- c(
  "london",
  "madrid",
  "amsterdam",
  "berlin",
  "rome",
  "venice",
  "vienna"
)

excluded_price_cities <- setdiff(all_cities, analytic_cities)

message(
  "Excluded because no usable listing-level price data were available: ",
  paste(excluded_price_cities, collapse = ", ")
)

cities <- analytic_cities

city_centres <- tibble::tribble(
  ~city,        ~lon,      ~lat,
  "london",     -0.1276,   51.5072,
  "madrid",     -3.7038,   40.4168,
  "amsterdam",   4.9041,   52.3676,
  "berlin",     13.4050,   52.5200,
  "rome",       12.4964,   41.9028,
  "venice",     12.3155,   45.4408,
  "vienna",     16.3738,   48.2082
)
```

Barcelona, Lisbon, and Paris are excluded because the downloaded files do not
contain usable nightly prices.

### Currency

London prices are converted from GBP using a fixed rate of 1.17 EUR.

```{r currency}
gbp_to_eur <- 1.17

currency_conversion <- tibble::tribble(
  ~city,        ~currency_to_eur,
  "london",     gbp_to_eur,
  "madrid",     1,
  "amsterdam",  1,
  "berlin",     1,
  "rome",       1,
  "venice",     1,
  "vienna",     1
)
```

## Input validation

```{r validate-inputs}
required_files <- tidyr::expand_grid(
  city = cities,
  file_type = c("listings", "neighbourhoods_geojson")
) %>%
  mutate(
    path = case_when(
      file_type == "listings" ~ file.path(data_dir, paste0("listings_", city, ".csv.gz")),
      file_type == "neighbourhoods_geojson" ~ file.path(
        data_dir,
        paste0("neighbourhoods_", city, ".geojson")
      )
    ),
    exists = file.exists(path)
  )

missing_files <- required_files %>%
  filter(!exists)

if (nrow(missing_files) > 0) {
  print(missing_files)
  stop("Some required files are missing or incorrectly named.")
}
```

## Helper functions

### Parsing

```{r parsing-helpers}
get_col <- function(df, colname, default = NA) {
  if (colname %in% names(df)) {
    df[[colname]]
  } else {
    rep(default, nrow(df))
  }
}

parse_tf <- function(x) {
  dplyr::case_when(
    x %in% c(TRUE, "t", "T", "true", "TRUE", "1", 1) ~ 1,
    x %in% c(FALSE, "f", "F", "false", "FALSE", "0", 0) ~ 0,
    TRUE ~ NA_real_
  )
}

parse_price <- function(x) {
  readr::parse_number(
    as.character(x),
    locale = readr::locale(grouping_mark = ",", decimal_mark = ".")
  )
}
```

### Scaling and labels

```{r transformation-helpers}
median_safe <- function(x) {
  m <- suppressWarnings(median(x, na.rm = TRUE))
  ifelse(is.finite(m), m, 0)
}

z <- function(x) {
  m <- mean(x, na.rm = TRUE)
  s <- sd(x, na.rm = TRUE)

  if (!is.finite(s) || s == 0) {
    return(rep(0, length(x)))
  }

  as.numeric((x - m) / s)
}

clean_neighbourhood <- function(x) {
  x %>%
    as.character() %>%
    stringr::str_squish()
}

```

## Listing data

### City reader

```{r read-listings}
read_city_listings <- function(city_name) {

  path <- file.path(data_dir, paste0("listings_", city_name, ".csv.gz"))

  message("Reading listings for: ", city_name)

  x <- readr::read_csv(
    path,
    show_col_types = FALSE,
    guess_max = 10000
  )

  bathroom_source <- if ("bathrooms" %in% names(x)) {
    get_col(x, "bathrooms")
  } else {
    get_col(x, "bathrooms_text")
  }

  tibble(
    city = city_name,
    listing_id = as.character(get_col(x, "id")),
    neighbourhood_original = clean_neighbourhood(get_col(x, "neighbourhood_cleansed")),
    latitude = as.numeric(get_col(x, "latitude")),
    longitude = as.numeric(get_col(x, "longitude")),

    price_raw = parse_price(get_col(x, "price")),

    room_type = stringr::str_squish(as.character(get_col(x, "room_type"))),
    room_type_clean = stringr::str_to_lower(
      stringr::str_squish(as.character(get_col(x, "room_type")))
    ),

    accommodates = as.numeric(get_col(x, "accommodates")),
    bedrooms = as.numeric(get_col(x, "bedrooms")),
    bathrooms = parse_price(bathroom_source),
    beds = as.numeric(get_col(x, "beds")),

    minimum_nights = as.numeric(get_col(x, "minimum_nights")),
    availability_365 = as.numeric(get_col(x, "availability_365")),

    number_of_reviews = as.numeric(get_col(x, "number_of_reviews")),
    review_scores_rating = as.numeric(get_col(x, "review_scores_rating")),

    host_is_superhost = parse_tf(get_col(x, "host_is_superhost")),
    instant_bookable = parse_tf(get_col(x, "instant_bookable"))
  )
}
```

### Combined listing table

```{r combine-listings}
listings_raw <- purrr::map_dfr(cities, read_city_listings) %>%
  left_join(currency_conversion, by = "city") %>%
  mutate(
    price_eur = price_raw * currency_to_eur
  ) %>%
  distinct(city, listing_id, .keep_all = TRUE)

raw_city_counts <- listings_raw %>%
  count(city, name = "raw_listings")

print(raw_city_counts)

listing_price_check <- listings_raw %>%
  group_by(city) %>%
  summarise(
    raw_listings = n(),
    usable_prices = sum(!is.na(price_eur) & price_eur > 0),
    missing_prices = sum(is.na(price_eur)),
    .groups = "drop"
  )

print(listing_price_check)

if (any(listing_price_check$usable_prices == 0)) {
  stop("At least one analytic city has no usable price values.")
}


```

## Neighbourhood boundaries

### Boundary reader

```{r read-neighbourhoods}
read_city_neighbourhood_shapes <- function(city_name) {

  path <- file.path(data_dir, paste0("neighbourhoods_", city_name, ".geojson"))

  message("Reading neighbourhood geometry for: ", city_name)

  g <- sf::st_read(path, quiet = TRUE)

  possible_name_cols <- c("neighbourhood", "neighbourhood_cleansed", "name")
  name_col <- possible_name_cols[possible_name_cols %in% names(g)][1]

  if (is.na(name_col)) {
    stop("No neighbourhood name column found in: ", path)
  }

  g <- g %>%
    mutate(
      city = city_name,
      geo_neighbourhood = clean_neighbourhood(.data[[name_col]])
    ) %>%
    filter(
      !is.na(geo_neighbourhood),
      geo_neighbourhood != ""
    ) %>%
    select(city, geo_neighbourhood, geometry)

  g <- g %>%
    sf::st_zm(drop = TRUE, what = "ZM") %>%
    sf::st_make_valid()

  g <- suppressWarnings(sf::st_collection_extract(g, "POLYGON"))
  g <- g[!sf::st_is_empty(g), ]

  sf::st_transform(g, 4326)
}
```

### Combined boundary table

```{r combine-neighbourhoods}
neighbourhood_shapes <- purrr::map_dfr(cities, read_city_neighbourhood_shapes)


```

## Neighbourhood geography

### Area and distance function

```{r neighbourhood-geography}
calculate_neighbourhood_metrics <- function(city_name) {

  g <- neighbourhood_shapes %>%
    filter(city == city_name)

  g_proj <- sf::st_transform(g, 3035)

  centre <- city_centres %>%
    filter(city == city_name)

  if (nrow(centre) != 1) {
    stop("City centre not found for: ", city_name)
  }

  centre_sf <- sf::st_as_sf(
    centre,
    coords = c("lon", "lat"),
    crs = 4326
  ) %>%
    sf::st_transform(3035)

  g_proj <- g_proj %>%
    mutate(
      area_km2_piece = as.numeric(sf::st_area(geometry)) / 1000000,
      dist_centre_km_piece = as.numeric(
        sf::st_distance(
          sf::st_point_on_surface(geometry),
          centre_sf[1, ]
        )
      ) / 1000
    )

  g_proj %>%
    sf::st_drop_geometry() %>%
    group_by(city, geo_neighbourhood) %>%
    summarise(
      area_km2 = sum(area_km2_piece, na.rm = TRUE),
      dist_centre_km = weighted.mean(
        dist_centre_km_piece,
        w = area_km2_piece,
        na.rm = TRUE
      ),
      .groups = "drop"
    ) %>%
    mutate(
      dist_centre_km = if_else(is.nan(dist_centre_km), NA_real_, dist_centre_km)
    )
}
```

### Area and distance table

```{r calculate-neighbourhood-geography}
neighbourhood_geo <- purrr::map_dfr(cities, calculate_neighbourhood_metrics)

print(
  neighbourhood_geo %>%
    group_by(city) %>%
    summarise(
      neighbourhoods_geo = n(),
      missing_area = sum(is.na(area_km2)),
      .groups = "drop"
    )
)


```

## Spatial matching

Listings are matched to the polygon that contains them. The nearest polygon is
used only when a point falls just outside all boundaries.

### Assignment function

```{r spatial-matching}
listings_raw <- listings_raw %>%
  mutate(row_id = row_number())

assign_listing_neighbourhoods <- function(city_name) {

  message("Spatially assigning listings for: ", city_name)

  pts_df <- listings_raw %>%
    filter(
      city == city_name,
      !is.na(longitude),
      !is.na(latitude),
      longitude >= -180,
      longitude <= 180,
      latitude >= -90,
      latitude <= 90
    )

  shapes <- neighbourhood_shapes %>%
    filter(city == city_name) %>%
    select(geo_neighbourhood)

  if (nrow(pts_df) == 0) {
    warning("No valid listing coordinates for: ", city_name)
    return(tibble(row_id = integer(), geo_neighbourhood = character()))
  }

  if (nrow(shapes) == 0) {
    warning("No neighbourhood shapes for: ", city_name)
    return(tibble(row_id = pts_df$row_id, geo_neighbourhood = NA_character_))
  }

  pts <- sf::st_as_sf(
    pts_df,
    coords = c("longitude", "latitude"),
    crs = 4326,
    remove = FALSE
  )

  joined <- sf::st_join(
    pts,
    shapes,
    join = sf::st_within,
    left = TRUE
  )

  joined_df <- joined %>%
    sf::st_drop_geometry() %>%
    group_by(row_id) %>%
    summarise(
      geo_neighbourhood = dplyr::first(
        geo_neighbourhood[!is.na(geo_neighbourhood)],
        default = NA_character_
      ),
      .groups = "drop"
    )

  unmatched <- joined_df %>%
    filter(is.na(geo_neighbourhood))

  if (nrow(unmatched) > 0) {

    unmatched_pts <- pts %>%
      filter(row_id %in% unmatched$row_id)

    nearest_id <- sf::st_nearest_feature(unmatched_pts, shapes)

    nearest_df <- tibble(
      row_id = unmatched_pts$row_id,
      geo_neighbourhood = shapes$geo_neighbourhood[nearest_id]
    )

    joined_df <- joined_df %>%
      filter(!row_id %in% unmatched$row_id) %>%
      bind_rows(nearest_df)
  }

  joined_df
}
```

### Apply the spatial match

```{r apply-spatial-matching}
listing_geo_lookup <- purrr::map_dfr(cities, assign_listing_neighbourhoods)

listings_raw_geo <- listings_raw %>%
  left_join(listing_geo_lookup, by = "row_id")

print(
  listings_raw_geo %>%
    group_by(city) %>%
    summarise(
      raw_listings = n(),
      assigned_geo_neighbourhood = sum(!is.na(geo_neighbourhood)),
      missing_geo_neighbourhood = sum(is.na(geo_neighbourhood)),
      .groups = "drop"
    )
)


```

## OpenStreetMap amenities

### Amenity groups

```{r osm-configuration}
osm_cache_dir <- file.path(output_dir, "osm_cache")
dir.create(osm_cache_dir, recursive = TRUE, showWarnings = FALSE)

osm_feature_groups <- list(
  amenity = c(
    "restaurant",
    "cafe",
    "bar",
    "pub",
    "fast_food",
    "food_court",
    "biergarten",
    "theatre",
    "cinema",
    "nightclub"
  ),
  tourism = c(
    "attraction",
    "museum",
    "gallery",
    "viewpoint",
    "artwork",
    "zoo",
    "aquarium"
  ),
  historic = c(
    "monument",
    "memorial",
    "castle",
    "archaeological_site"
  )
)

empty_poi_sf <- function() {
  sf::st_sf(
    city = character(),
    osm_id = character(),
    osm_key = character(),
    osm_value = character(),
    osm_layer = character(),
    geometry = sf::st_sfc(crs = 4326)
  )
}
```

### Geometry extraction

```{r extract-osm-layer}
extract_osm_layer <- function(osm_result, layer_name, city_name, osm_key) {

  layer <- osm_result[[layer_name]]

  if (is.null(layer) || nrow(layer) == 0) {
    return(empty_poi_sf())
  }

  layer <- sf::st_transform(layer, 4326)

  if (layer_name != "osm_points") {
    layer <- suppressWarnings(sf::st_make_valid(layer))
    sf::st_geometry(layer) <- suppressWarnings(
      sf::st_point_on_surface(sf::st_geometry(layer))
    )
  }

  if (!"osm_id" %in% names(layer)) {
    layer$osm_id <- NA_character_
  }

  if (osm_key %in% names(layer)) {
    layer$osm_value <- as.character(layer[[osm_key]])
  } else {
    layer$osm_value <- NA_character_
  }

  layer$city <- city_name
  layer$osm_key <- osm_key
  layer$osm_layer <- layer_name

  layer %>%
    select(city, osm_id, osm_key, osm_value, osm_layer, geometry)
}
```

### OSM query

```{r download-osm-group}
download_city_osm_group <- function(city_name, osm_key, osm_values) {

  message("Downloading OSM group for ", city_name, ": ", osm_key)

  city_shape <- neighbourhood_shapes %>%
    filter(city == city_name) %>%
    sf::st_transform(4326)

  bbox <- sf::st_bbox(city_shape)

  q <- osmdata::opq(
    bbox = unname(c(
      bbox["xmin"],
      bbox["ymin"],
      bbox["xmax"],
      bbox["ymax"]
    )),
    timeout = 180
  ) %>%
    osmdata::add_osm_feature(
      key = osm_key,
      value = osm_values,
      value_exact = TRUE
    )

  osm_result <- tryCatch(
    osmdata::osmdata_sf(q),
    error = function(e) {
      warning(
        "OSM download failed for ",
        city_name,
        " / ",
        osm_key,
        ": ",
        conditionMessage(e)
      )
      return(NULL)
    }
  )

  if (is.null(osm_result)) {
    return(empty_poi_sf())
  }

  pois <- bind_rows(
    extract_osm_layer(osm_result, "osm_points", city_name, osm_key),
    extract_osm_layer(osm_result, "osm_polygons", city_name, osm_key),
    extract_osm_layer(osm_result, "osm_multipolygons", city_name, osm_key)
  )

  if (nrow(pois) == 0) {
    return(empty_poi_sf())
  }

  pois
}
```

### City cache

```{r download-city-osm}
download_city_osm_pois <- function(city_name) {

  cache_path <- file.path(osm_cache_dir, paste0("osm_tourist_pois_", city_name, ".rds"))

  if (file.exists(cache_path)) {
    message("Reading cached OSM POIs for: ", city_name)
    return(readRDS(cache_path))
  }

  message("Downloading OSM tourist POIs for: ", city_name)

  city_pois <- purrr::imap_dfr(
    osm_feature_groups,
    ~download_city_osm_group(
      city_name = city_name,
      osm_key = .y,
      osm_values = .x
    )
  )

  if (nrow(city_pois) > 0) {
    city_pois <- city_pois %>%
      mutate(
        osm_unique_id = case_when(
          !is.na(osm_id) & osm_id != "" ~ paste(osm_key, osm_id, sep = ":"),
          TRUE ~ paste(osm_key, osm_value, row_number(), sep = ":")
        )
      ) %>%
      distinct(osm_unique_id, .keep_all = TRUE)
  }

  saveRDS(city_pois, cache_path)

  city_pois
}
```

### Combined amenity table

```{r collect-osm-amenities}
osm_tourist_pois <- purrr::map_dfr(cities, download_city_osm_pois)

print(
  osm_tourist_pois %>%
    sf::st_drop_geometry() %>%
    count(city, name = "tourist_pois")
)


```

Amenities include food and nightlife venues, cultural attractions, and
historic visitor sites. Polygon features are represented by a point on their
surface before they are assigned to a neighbourhood.

### Neighbourhood counts

```{r neighbourhood-amenities}
count_pois_by_city_neighbourhood <- function(city_name) {

  message("Counting OSM POIs by neighbourhood for: ", city_name)

  city_neighbourhoods <- neighbourhood_shapes %>%
    filter(city == city_name) %>%
    sf::st_transform(4326) %>%
    select(geo_neighbourhood)

  all_neighbourhoods <- city_neighbourhoods %>%
    sf::st_drop_geometry() %>%
    distinct(geo_neighbourhood) %>%
    mutate(city = city_name)

  city_pois <- osm_tourist_pois %>%
    filter(city == city_name) %>%
    sf::st_transform(4326)

  if (nrow(city_pois) == 0) {
    return(
      all_neighbourhoods %>%
        mutate(tourist_amenities = 0L)
    )
  }

  joined <- sf::st_join(
    city_pois,
    city_neighbourhoods,
    join = sf::st_intersects,
    left = FALSE
  )

  if (nrow(joined) == 0) {
    return(
      all_neighbourhoods %>%
        mutate(tourist_amenities = 0L)
    )
  }

  counts <- joined %>%
    sf::st_drop_geometry() %>%
    distinct(osm_unique_id, geo_neighbourhood, .keep_all = TRUE) %>%
    count(city, geo_neighbourhood, name = "tourist_amenities")

  all_neighbourhoods %>%
    left_join(counts, by = c("city", "geo_neighbourhood")) %>%
    mutate(
      tourist_amenities = replace_na(tourist_amenities, 0L)
    )
}
```

### Amenity density

```{r calculate-neighbourhood-amenities}
tourist_amenity_counts <- purrr::map_dfr(
  cities,
  count_pois_by_city_neighbourhood
)

tourist_amenity_context <- tourist_amenity_counts %>%
  left_join(
    neighbourhood_geo,
    by = c("city", "geo_neighbourhood"),
    relationship = "many-to-one"
  ) %>%
  mutate(
    tourist_amenity_density = tourist_amenities / area_km2,
    log_tourist_amenity_density = log1p(tourist_amenity_density)
  ) %>%
  select(
    city,
    geo_neighbourhood,
    tourist_amenities,
    tourist_amenity_density,
    log_tourist_amenity_density
  )

print(
  tourist_amenity_context %>%
    group_by(city) %>%
    summarise(
      neighbourhoods = n(),
      total_tourist_amenities = sum(tourist_amenities, na.rm = TRUE),
      median_tourist_amenity_density = median(tourist_amenity_density, na.rm = TRUE),
      .groups = "drop"
    )
)

readr::write_csv(
  tourist_amenity_context,
  file.path(output_dir, "tourist_amenity_context.csv")
)


```

## Neighbourhood context

### Listing-level ingredients

```{r neighbourhood-context}
context_base <- listings_raw_geo %>%
  filter(
    !is.na(city),
    !is.na(listing_id),
    !is.na(geo_neighbourhood),
    geo_neighbourhood != ""
  ) %>%
  mutate(
    entire_home_context = if_else(room_type_clean == "entire home/apt", 1, 0),
    number_of_reviews_context = replace_na(number_of_reviews, 0),
    availability_context = replace_na(availability_365, 0)
  )
```

### Neighbourhood aggregates

```{r aggregate-neighbourhood-context}
neighbourhood_context <- context_base %>%
  group_by(city, geo_neighbourhood) %>%
  summarise(
    n_listings_all = n(),
    n_entire_home_all = sum(entire_home_context, na.rm = TRUE),
    sum_reviews_all = sum(number_of_reviews_context, na.rm = TRUE),
    sum_availability_all = sum(availability_context, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(
    neighbourhood_geo,
    by = c("city", "geo_neighbourhood"),
    relationship = "many-to-one"
  ) %>%
  left_join(
    tourist_amenity_context,
    by = c("city", "geo_neighbourhood"),
    relationship = "many-to-one"
  ) %>%
  mutate(
    tourist_amenities = replace_na(tourist_amenities, 0L),
    tourist_amenity_density = replace_na(tourist_amenity_density, 0),
    log_tourist_amenity_density = replace_na(log_tourist_amenity_density, 0),

    listing_density_all = n_listings_all / area_km2,
    share_entire_home_all = n_entire_home_all / n_listings_all,
    mean_reviews_all = sum_reviews_all / n_listings_all,
    mean_availability_all = sum_availability_all / n_listings_all
  )

print(
  neighbourhood_context %>%
    group_by(city) %>%
    summarise(
      neighbourhoods = n(),
      missing_area = sum(is.na(area_km2)),
      missing_tourist_amenity_density = sum(is.na(tourist_amenity_density)),
      .groups = "drop"
    )
)


```

## Analysis sample

### Listing-level cleaning

```{r clean-listings}
model_base <- listings_raw_geo %>%
  mutate(
    entire_home = case_when(
      room_type_clean == "entire home/apt" ~ 1,
      room_type_clean == "private room" ~ 0,
      TRUE ~ NA_real_
    )
  ) %>%
  filter(
    !is.na(price_eur),
    price_eur > 0,
    !is.na(geo_neighbourhood),
    geo_neighbourhood != "",
    !is.na(city),
    !is.na(entire_home)
  ) %>%
  group_by(city) %>%
  mutate(
    price_p01 = quantile(price_eur, 0.01, na.rm = TRUE),
    price_p99 = quantile(price_eur, 0.99, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(
    price_eur >= price_p01,
    price_eur <= price_p99
  ) %>%
  select(-price_p01, -price_p99) %>%
  group_by(city) %>%
  mutate(
    accommodates = replace_na(accommodates, median_safe(accommodates)),
    bedrooms = replace_na(bedrooms, median_safe(bedrooms)),
    bathrooms = replace_na(bathrooms, median_safe(bathrooms)),
    beds = replace_na(beds, median_safe(beds)),
    minimum_nights = replace_na(minimum_nights, median_safe(minimum_nights)),
    availability_365 = replace_na(availability_365, median_safe(availability_365)),

    review_missing = if_else(is.na(review_scores_rating), 1, 0),
    review_scores_rating = replace_na(
      review_scores_rating,
      median_safe(review_scores_rating)
    ),

    number_of_reviews = replace_na(number_of_reviews, 0),
    host_is_superhost = replace_na(host_is_superhost, 0),
    instant_bookable = replace_na(instant_bookable, 0)
  ) %>%
  ungroup() %>%
  mutate(
    log_price = log(price_eur),
    log_reviews = log1p(number_of_reviews),
    log_minimum_nights = log1p(minimum_nights)
  )


```

### Leave-one-out context

```{r leave-one-out-context}
model_data <- model_base %>%
  left_join(
    neighbourhood_context,
    by = c("city", "geo_neighbourhood"),
    relationship = "many-to-one"
  ) %>%
  mutate(
    n_other_listings = n_listings_all - 1,

    density_loo = if_else(
      n_other_listings > 0,
      n_other_listings / area_km2,
      NA_real_
    ),

    share_entire_home_loo = if_else(
      n_other_listings > 0,
      (n_entire_home_all - entire_home) / n_other_listings,
      NA_real_
    ),

    mean_reviews_loo = if_else(
      n_other_listings > 0,
      (sum_reviews_all - number_of_reviews) / n_other_listings,
      NA_real_
    ),

    mean_availability_loo = if_else(
      n_other_listings > 0,
      (sum_availability_all - availability_365) / n_other_listings,
      NA_real_
    )
  ) %>%
  filter(
    n_listings_all >= 20,
    !is.na(area_km2),
    area_km2 > 0,
    !is.na(density_loo),
    !is.na(dist_centre_km)
  )
```

### Model variables

Distance to the city centre is standardised within each city so that the
measure remains comparable across differently sized cities.

```{r build-model-data}
model_data <- model_data %>%
  mutate(
    city = factor(city),
    neighbourhood = geo_neighbourhood,
    neighbourhood_id = factor(paste(city, geo_neighbourhood, sep = "__")),

    accommodates_z = z(accommodates),
    bedrooms_z = z(bedrooms),
    bathrooms_z = z(bathrooms),
    beds_z = z(beds),
    rating_z = z(review_scores_rating),
    log_reviews_z = z(log_reviews),
    log_minimum_nights_z = z(log_minimum_nights),
    availability_365_z = z(availability_365),

    density_z = z(log1p(density_loo)),
    share_entire_home_z = z(share_entire_home_loo),
    mean_reviews_z = z(log1p(mean_reviews_loo)),
    mean_availability_z = z(mean_availability_loo),
    tourist_amenity_density_z = z(log_tourist_amenity_density)
  ) %>%
  group_by(city) %>%
  mutate(
    dist_centre_z = z(dist_centre_km)
  ) %>%
  ungroup() %>%
  drop_na(
    log_price,
    entire_home,
    accommodates_z,
    bedrooms_z,
    bathrooms_z,
    rating_z,
    log_reviews_z,
    log_minimum_nights_z,
    availability_365_z,
    density_z,
    share_entire_home_z,
    mean_reviews_z,
    mean_availability_z,
    tourist_amenity_density_z,
    dist_centre_z,
    city,
    neighbourhood_id
  )


```

## Diagnostics

### Sample flow

```{r diagnostic-counts}
diagnostic_raw <- listings_raw %>%
  count(city, name = "raw")

diagnostic_valid_coords <- listings_raw %>%
  filter(
    !is.na(longitude),
    !is.na(latitude),
    longitude >= -180,
    longitude <= 180,
    latitude >= -90,
    latitude <= 90
  ) %>%
  count(city, name = "valid_coords")

diagnostic_geo_assigned <- listings_raw_geo %>%
  filter(!is.na(geo_neighbourhood)) %>%
  count(city, name = "geo_assigned")

diagnostic_context <- neighbourhood_context %>%
  filter(!is.na(area_km2), area_km2 > 0) %>%
  group_by(city) %>%
  summarise(
    context_neighbourhoods = n(),
    context_listings = sum(n_listings_all, na.rm = TRUE),
    .groups = "drop"
  )

diagnostic_model_base <- model_base %>%
  count(city, name = "model_base")

diagnostic_model_data <- model_data %>%
  count(city, name = "model_data")
```

```{r sample-flow}
city_loss_diagnostic <- diagnostic_raw %>%
  full_join(diagnostic_valid_coords, by = "city") %>%
  full_join(diagnostic_geo_assigned, by = "city") %>%
  full_join(diagnostic_context, by = "city") %>%
  full_join(diagnostic_model_base, by = "city") %>%
  full_join(diagnostic_model_data, by = "city") %>%
  mutate(across(where(is.numeric), ~replace_na(.x, 0))) %>%
  arrange(city)

print(city_loss_diagnostic)

readr::write_csv(
  city_loss_diagnostic,
  file.path(output_dir, "city_loss_diagnostic.csv")
)
```

### Retention by city

```{r city-retention}
model_city_counts <- model_data %>%
  count(city, name = "model_listings")

city_retention <- raw_city_counts %>%
  left_join(model_city_counts, by = "city") %>%
  mutate(
    model_listings = replace_na(model_listings, 0),
    retained_share = model_listings / raw_listings
  )

print(city_retention)

dropped_cities <- setdiff(cities, unique(as.character(model_data$city)))

if (length(dropped_cities) > 0) {
  warning(
    "These analytic cities are absent from model_data: ",
    paste(dropped_cities, collapse = ", ")
  )
}

readr::write_csv(
  model_data,
  file.path(output_dir, "airbnb_multilevel_model_data_osm.csv")
)


```

## Descriptive statistics

### City summary

```{r city-summary}
city_summary <- model_data %>%
  group_by(city) %>%
  summarise(
    listings = n(),
    neighbourhoods = n_distinct(neighbourhood_id),
    median_price_eur = median(price_eur, na.rm = TRUE),
    mean_log_price = mean(log_price, na.rm = TRUE),
    share_entire_home = mean(entire_home, na.rm = TRUE),
    median_tourist_amenity_density = median(tourist_amenity_density, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(median_price_eur))

print(city_summary)
```

### Neighbourhood summary

```{r neighbourhood-summary}
neighbourhood_summary <- model_data %>%
  group_by(city, neighbourhood) %>%
  summarise(
    listings = n(),
    median_price_eur = median(price_eur, na.rm = TRUE),
    density = first(density_loo),
    share_entire_home = first(share_entire_home_loo),
    dist_centre_km = first(dist_centre_km),
    tourist_amenities = first(tourist_amenities),
    tourist_amenity_density = first(tourist_amenity_density),
    .groups = "drop"
  )

print(head(neighbourhood_summary, 30))


```

## Exploratory plots

### City prices

```{r city-price-plot, fig.width=8, fig.height=5}
city_price_plot_data <- model_data %>%
  group_by(city) %>%
  summarise(
    median_price = median(price_eur, na.rm = TRUE),
    p25 = quantile(price_eur, 0.25, na.rm = TRUE),
    p75 = quantile(price_eur, 0.75, na.rm = TRUE),
    .groups = "drop"
  )

plot_city_prices <- ggplot(
  city_price_plot_data,
  aes(y = forcats::fct_reorder(city, median_price), x = median_price)
) +
  geom_errorbarh(aes(xmin = p25, xmax = p75), height = 0.18) +
  geom_point(size = 2) +
  labs(
    title = "Median Airbnb price by city",
    subtitle = "Points show medians; bars show interquartile ranges",
    x = "Nightly price, approximately EUR",
    y = "City"
  ) +
  theme_minimal(base_size = 13)

print(plot_city_prices)

ggsave(
  filename = file.path(output_dir, "plot_city_prices_clean.png"),
  plot = plot_city_prices,
  width = 8,
  height = 5,
  dpi = 300
)
```

### Listing density

```{r density-price-plot, fig.width=10, fig.height=6}
plot_density_price <- ggplot(
  neighbourhood_summary,
  aes(x = log1p(density), y = median_price_eur)
) +
  geom_point(alpha = 0.45, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  facet_wrap(~ city, scales = "free_y") +
  labs(
    title = "Neighbourhood Airbnb density and median nightly price",
    x = "Log neighbourhood Airbnb density",
    y = "Median nightly price, approximately EUR"
  ) +
  theme_minimal(base_size = 12)

print(plot_density_price)

ggsave(
  filename = file.path(output_dir, "plot_density_price_by_city.png"),
  plot = plot_density_price,
  width = 10,
  height = 6,
  dpi = 300
)
```

### Tourist amenity density

```{r tourist-price-plot, fig.width=10, fig.height=6}
plot_tourist_price <- ggplot(
  neighbourhood_summary,
  aes(x = log1p(tourist_amenity_density), y = median_price_eur)
) +
  geom_point(alpha = 0.45, size = 1.5) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8) +
  facet_wrap(~ city, scales = "free_y") +
  labs(
    title = "Tourist-amenity density and median Airbnb price",
    x = "Log tourist-amenity density",
    y = "Median nightly price, approximately EUR"
  ) +
  theme_minimal(base_size = 12)

print(plot_tourist_price)

ggsave(
  filename = file.path(output_dir, "plot_tourist_amenity_price_by_city.png"),
  plot = plot_tourist_price,
  width = 10,
  height = 6,
  dpi = 300
)


```

## Multilevel models

### Optimizer

```{r model-control}
ctrl <- lmerControl(
  optimizer = "bobyqa",
  optCtrl = list(maxfun = 200000)
)
```

### Empty model

```{r empty-model}
m0 <- lmer(
  log_price ~ 1 +
    (1 | city) +
    (1 | neighbourhood_id),
  data = model_data,
  REML = FALSE,
  control = ctrl
)
```

### Listing controls

```{r listing-model}
m1 <- lmer(
  log_price ~
    entire_home +
    accommodates_z +
    bedrooms_z +
    bathrooms_z +
    rating_z +
    review_missing +
    log_reviews_z +
    host_is_superhost +
    instant_bookable +
    log_minimum_nights_z +
    availability_365_z +
    (1 | city) +
    (1 | neighbourhood_id),
  data = model_data,
  REML = FALSE,
  control = ctrl
)
```

### Neighbourhood context

```{r context-model}
m2 <- lmer(
  log_price ~
    entire_home +
    accommodates_z +
    bedrooms_z +
    bathrooms_z +
    rating_z +
    review_missing +
    log_reviews_z +
    host_is_superhost +
    instant_bookable +
    log_minimum_nights_z +
    availability_365_z +
    density_z +
    share_entire_home_z +
    mean_reviews_z +
    mean_availability_z +
    tourist_amenity_density_z +
    dist_centre_z +
    (1 | city) +
    (1 | neighbourhood_id),
  data = model_data,
  REML = FALSE,
  control = ctrl
)
```

### Random slope

```{r random-slope-model}
m3 <- lmer(
  log_price ~
    entire_home +
    accommodates_z +
    bedrooms_z +
    bathrooms_z +
    rating_z +
    review_missing +
    log_reviews_z +
    host_is_superhost +
    instant_bookable +
    log_minimum_nights_z +
    availability_365_z +
    density_z +
    share_entire_home_z +
    mean_reviews_z +
    mean_availability_z +
    tourist_amenity_density_z +
    dist_centre_z +
    (1 | city) +
    (1 + entire_home || neighbourhood_id),
  data = model_data,
  REML = FALSE,
  control = ctrl
)
```

### Cross-level interaction

```{r interaction-model}
m4 <- lmer(
  log_price ~
    entire_home * density_z +
    accommodates_z +
    bedrooms_z +
    bathrooms_z +
    rating_z +
    review_missing +
    log_reviews_z +
    host_is_superhost +
    instant_bookable +
    log_minimum_nights_z +
    availability_365_z +
    share_entire_home_z +
    mean_reviews_z +
    mean_availability_z +
    tourist_amenity_density_z +
    dist_centre_z +
    (1 | city) +
    (1 | neighbourhood_id),
  data = model_data,
  REML = FALSE,
  control = ctrl
)
```

### Final ML specification

```{r final-ml-model}
m5 <- lmer(
  log_price ~
    entire_home * density_z +
    accommodates_z +
    bedrooms_z +
    bathrooms_z +
    rating_z +
    review_missing +
    log_reviews_z +
    host_is_superhost +
    instant_bookable +
    log_minimum_nights_z +
    availability_365_z +
    share_entire_home_z +
    mean_reviews_z +
    mean_availability_z +
    tourist_amenity_density_z +
    dist_centre_z +
    (1 | city) +
    (1 + entire_home || neighbourhood_id),
  data = model_data,
  REML = FALSE,
  control = ctrl
)


```

### Model comparison

```{r compare-models}
summary(m0)
summary(m1)
summary(m2)
summary(m3)
summary(m4)
summary(m5)

anova(m0, m1, m2, m4)
anova(m2, m3)
anova(m3, m5)

singularity_check <- tibble(
  model = c("m0", "m1", "m2", "m3", "m4", "m5"),
  singular = c(
    isSingular(m0),
    isSingular(m1),
    isSingular(m2),
    isSingular(m3),
    isSingular(m4),
    isSingular(m5)
  )
)

print(singularity_check)


```

### Variance decomposition

```{r variance-decomposition}
variance_partition <- function(model) {
  as.data.frame(VarCorr(model)) %>%
    filter(is.na(var2)) %>%
    transmute(
      level = case_when(
        grp == "Residual" ~ "Residual",
        TRUE ~ paste(grp, var1)
      ),
      variance = vcov
    ) %>%
    mutate(
      total_variance = sum(variance),
      proportion = variance / total_variance
    )
}

variance_reduction <- function(base_model, new_model) {
  base <- variance_partition(base_model) %>%
    select(level, base_variance = variance)

  new <- variance_partition(new_model) %>%
    select(level, new_variance = variance)

  base %>%
    inner_join(new, by = "level") %>%
    mutate(
      reduction = 1 - (new_variance / base_variance)
    )
}

print(variance_partition(m0))
print(variance_partition(m1))
print(variance_partition(m2))
print(variance_partition(m4))

print(variance_reduction(m0, m1))
print(variance_reduction(m1, m2))
print(variance_reduction(m2, m4))


```

### Model table

```{r model-table}
models <- list(
  "M0 Empty" = m0,
  "M1 Listing controls" = m1,
  "M2 Context + OSM" = m2,
  "M3 Random slope + OSM" = m3,
  "M4 Interaction + OSM" = m4,
  "M5 Final + OSM" = m5
)

modelsummary(
  models,
  statistic = "({std.error})",
  stars = TRUE,
  output = "markdown"
)

modelsummary(
  models,
  statistic = "({std.error})",
  stars = TRUE,
  output = file.path(output_dir, "airbnb_model_table_osm.html")
)


```

### Interaction effect

```{r interaction-plot, fig.width=8, fig.height=5}
interaction_predictions <- ggeffects::ggpredict(
  m5,
  terms = c("density_z [-2,-1,0,1,2]", "entire_home [0,1]")
)

interaction_df <- as.data.frame(interaction_predictions) %>%
  mutate(
    listing_type = case_when(
      as.character(group) %in% c("0", "0.0") ~ "Private room",
      as.character(group) %in% c("1", "1.0") ~ "Entire home/apartment",
      TRUE ~ as.character(group)
    )
  )

plot_interaction <- ggplot(
  interaction_df,
  aes(x = x, y = exp(predicted), linetype = listing_type)
) +
  geom_line(linewidth = 1) +
  geom_ribbon(
    aes(
      ymin = exp(conf.low),
      ymax = exp(conf.high),
      group = listing_type
    ),
    alpha = 0.15,
    colour = NA
  ) +
  labs(
    title = "Predicted Airbnb price by listing type and neighbourhood density",
    subtitle = "Predictions from final OSM-enhanced multilevel model",
    x = "Neighbourhood Airbnb density, standardized",
    y = "Predicted nightly price, approximately EUR",
    linetype = "Listing type"
  ) +
  theme_minimal(base_size = 13)

print(plot_interaction)

ggsave(
  filename = file.path(output_dir, "plot_interaction_final_osm.png"),
  plot = plot_interaction,
  width = 8,
  height = 5,
  dpi = 300
)


```

### City-level effects

```{r city-effects, fig.width=8, fig.height=5}
city_random_effects <- ranef(m0)$city %>%
  rownames_to_column("city") %>%
  rename(random_intercept = `(Intercept)`)

plot_city_random_intercepts <- ggplot(
  city_random_effects,
  aes(x = random_intercept, y = forcats::fct_reorder(city, random_intercept))
) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 2) +
  labs(
    title = "City-level random intercepts from the empty model",
    x = "City random intercept",
    y = "City"
  ) +
  theme_minimal(base_size = 13)

print(plot_city_random_intercepts)

ggsave(
  filename = file.path(output_dir, "plot_city_random_intercepts.png"),
  plot = plot_city_random_intercepts,
  width = 8,
  height = 5,
  dpi = 300
)


```

### Final model

```{r final-model}
m_final_reml <- update(m5, REML = TRUE)

summary(m_final_reml)
variance_partition(m_final_reml)

saveRDS(
  m_final_reml,
  file.path(output_dir, "m_final_reml_osm.rds")
)

fixed_effects <- fixef(m_final_reml)

effect_interpretation <- tibble(
  term = names(fixed_effects),
  estimate_log_points = as.numeric(fixed_effects),
  approximate_percent_change = 100 * (exp(estimate_log_points) - 1)
)

print(effect_interpretation)

readr::write_csv(
  effect_interpretation,
  file.path(output_dir, "effect_interpretation_percent_changes_osm.csv")
)


```

## Export

The compact output tables are saved separately from the rendered report.

```{r export-results}
readr::write_csv(city_summary, file.path(output_dir, "city_summary_osm.csv"))
readr::write_csv(neighbourhood_summary, file.path(output_dir, "neighbourhood_summary_osm.csv"))
readr::write_csv(city_retention, file.path(output_dir, "city_retention_osm.csv"))
readr::write_csv(singularity_check, file.path(output_dir, "singularity_check_osm.csv"))

message("Analysis outputs saved to: ", output_dir)
```

## Publication figures

### Shared style

```{r publication-plot-setup}
plot_dir <- file.path(output_dir, "figures")
dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

city_order <- c(
  "amsterdam",
  "london",
  "venice",
  "rome",
  "madrid",
  "berlin",
  "vienna"
)

city_labels <- c(
  amsterdam = "Amsterdam",
  london = "London",
  venice = "Venice",
  rome = "Rome",
  madrid = "Madrid",
  berlin = "Berlin",
  vienna = "Vienna"
)

publication_plot_data <- neighbourhood_summary %>%
  mutate(city = as.character(city)) %>%
  filter(city %in% city_order)

theme_city_minimal <- function(base_size = 10.5) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(
        face = "bold",
        size = base_size + 1.2,
        hjust = 0.5,
        margin = margin(b = 6)
      ),
      axis.title = element_blank(),
      axis.text = element_text(size = base_size, colour = "grey20"),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey87", linewidth = 0.35),
      plot.margin = margin(t = 6, r = 8, b = 6, l = 8)
    )
}

line_blue <- "#3568A8"
line_red <- "#9B3A3A"
point_colour <- "grey55"
band_colour <- "grey86"
```

### City panels

```{r city-panel-function}
make_city_panel <- function(
  data,
  city_name,
  x_var,
  line_colour,
  x_limits = c(0, 8),
  y_limits = c(0, 310),
  x_breaks = c(0, 2, 4, 6, 8),
  y_breaks = c(0, 100, 200, 300)
) {

  d <- data %>%
    filter(city == city_name)

  ggplot(d, aes(x = .data[[x_var]], y = median_price_eur)) +
    geom_point(
      size = 2.2,
      alpha = 0.55,
      colour = point_colour
    ) +
    geom_smooth(
      method = "lm",
      formula = y ~ x,
      se = TRUE,
      linewidth = 0.9,
      colour = line_colour,
      fill = band_colour
    ) +
    scale_x_continuous(
      limits = x_limits,
      breaks = x_breaks,
      expand = expansion(mult = c(0.03, 0.03))
    ) +
    scale_y_continuous(
      limits = y_limits,
      breaks = y_breaks,
      expand = expansion(mult = c(0.03, 0.03))
    ) +
    labs(title = city_labels[[city_name]]) +
    theme_city_minimal()
}
```

### Figure layout

```{r horizontal-grid-function}
make_horizontal_grid <- function(
  plot_list,
  x_label,
  y_label,
  save_path,
  width = 14,
  height = 8.2
) {

  blank_panel <- ggplot() + theme_void()

  top_row <- cowplot::plot_grid(
    plotlist = plot_list[1:4],
    ncol = 4,
    align = "hv"
  )

  bottom_row <- cowplot::plot_grid(
    plotlist = c(plot_list[5:7], list(blank_panel)),
    ncol = 4,
    align = "hv"
  )

  full_grid <- cowplot::plot_grid(
    top_row,
    bottom_row,
    ncol = 1,
    rel_heights = c(1, 1)
  )

  final_plot <- cowplot::ggdraw() +
    cowplot::draw_plot(full_grid, x = 0.055, y = 0.08, width = 0.93, height = 0.88) +
    cowplot::draw_label(
      x_label,
      x = 0.53,
      y = 0.02,
      size = 13,
      hjust = 0.5
    ) +
    cowplot::draw_label(
      y_label,
      x = 0.02,
      y = 0.50,
      size = 13,
      angle = 90,
      hjust = 0.5
    )

  print(final_plot)

  ggsave(
    filename = save_path,
    plot = final_plot,
    width = width,
    height = height,
    dpi = 300,
    bg = "white"
  )

  invisible(final_plot)
}
```

### Tourist amenities

```{r tourist-amenity-figure, fig.width=14, fig.height=8.2}

tourist_plot_data <- publication_plot_data %>%
  mutate(log_tourist_amenity_density = log1p(tourist_amenity_density))

tourist_panels <- lapply(
  city_order,
  function(city_name) {
    make_city_panel(
      data = tourist_plot_data,
      city_name = city_name,
      x_var = "log_tourist_amenity_density",
      line_colour = line_blue
    )
  }
)

plot_tourist_horizontal <- make_horizontal_grid(
  plot_list = tourist_panels,
  x_label = "Log tourist-amenity density",
  y_label = "Median nightly price, approximately EUR",
  save_path = file.path(plot_dir, "figure_01_tourist_amenities_horizontal.png"),
  width = 14,
  height = 8.2
)
```

### Airbnb density

```{r airbnb-density-figure, fig.width=14, fig.height=8.2}

density_plot_data <- publication_plot_data %>%
  mutate(log_airbnb_density = log1p(density))

density_panels <- lapply(
  city_order,
  function(city_name) {
    make_city_panel(
      data = density_plot_data,
      city_name = city_name,
      x_var = "log_airbnb_density",
      line_colour = line_red
    )
  }
)

plot_density_horizontal <- make_horizontal_grid(
  plot_list = density_panels,
  x_label = "Log Airbnb listing density",
  y_label = "Median nightly price, approximately EUR",
  save_path = file.path(plot_dir, "figure_02_airbnb_density_horizontal.png"),
  width = 14,
  height = 8.2
)

message("Publication figures saved to: ", plot_dir)
```

## Package citations

```{r package-citations}
citation("sf")
citation("osmdata")
```
